---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# journeyer

<!-- badges: start -->
[![R-CMD-check](https://github.com/inpowell/journeyer/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/inpowell/journeyer/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `journeyer` is to provide flexible strategies to follow
patient journeys through administrative data from NSW Health. The
simplest administrative patient journey is the *period of hospital care*
(POHC), which aims to capture all episodes of care in a hospitalisation
from the time a patient is admitted to when they are formally
discharged.

## Installation

You can install the development version of `journeyer` from GitHub using the
`remotes` package:

```r
remotes::install_github('inpowell/journeyer')
```

## Example

The following presents an example of how the `journeyer` package should
be used. This uses an artificial dataset that mimics some of the
patterns found in the NSW APDC. The toy APDC dataset contains 21,892
records and 9 variables. `RL_ID` is a unique record identifier, while
`PPN` is a person identifier. The facility columns all represent the
first mentioned facility for a person by A, the second by B, and so on.
The source of referral and mode of separation columns use codes as
defined in the [NSW APDC data
dictionary](https://www.cherel.org.au/data-dictionaries).

```{r}
library(dplyr) # For data manipulation
library(data.table)

library(journeyer)

data(toy_apdc)
head(toy_apdc, 10L)
```

Using this data we have read in, we now find links between these
episodes. To start, we need to join the dataset to itself using the PPN
column, and filter out pairs relating to the same record. By convention,
the suffixes to resolve name collisions should be `_parent` and
`_child`.

```{r}
pairs <- inner_join(toy_apdc, toy_apdc, by = 'PPN',
                    suffix = c('_parent', '_child'),
                    relationship = 'many-to-many') |>
  filter(RL_ID_parent != RL_ID_child)
```

Now that the pairs are lined up side-by-side, we can use the default
predicates for the NSW APDC as they are present in the package. For
convenience, we keep only the identifier columns.

```{r}
links <- pairs |>
  find_links(!!!default_predicates$NSW$APDC_APDC) |>
  select(RL_ID_parent, RL_ID_child)

head(links)
```

Now, we load on a new `journey` column onto the toy data, which requires
the toy data to be in `data.table` format.

```{r}
setDT(toy_apdc)
collect_journeys(toy_apdc, links, identifier = 'RL_ID')

head(toy_apdc, 10L)
```

Now, we can count the number of POHCs in the toy data:

```{r}
toy_apdc[, uniqueN(journey)]
```

We can find the start and end of each POHC, and count the number of
episodes in each POHC:

```{r}
head(toy_apdc[, .(start = min(episode_start), end = max(episode_end), Neps = .N), by = 'journey'])
```

A researcher may also calculate summary information for each POHC, such
as first primary procedure, the presence of certain diagnosis codes, and
demographic information. However, this data manipulation is beyond the
scope of this README.

Most POHCs have just one episode, which is similar to the NSW APDC:

```{r}
toy_apdc[, .("Episodes in POHC" = .N), by = 'journey'][, .(POHCs = .N), keyby = 'Episodes in POHC']
```

## Using different predicates

The above example uses default predicates which are coded in the
`journeyer` package. However, different predicates can be used! First,
let us consider a simple example where episodes are considered linked if
and only if they overlap. We could find these links as follows.

```{r}
pairs <- inner_join(toy_apdc, toy_apdc, by = 'PPN',
                    suffix = c('_parent', '_child'),
                    relationship = 'many-to-many') |>
  filter(RL_ID_parent != RL_ID_child)

links <- pairs |>
  find_links(episode_start_parent <= episode_start_child & episode_start_child <= episode_end_parent) |>
  select(RL_ID_parent, RL_ID_child)

collect_journeys(toy_apdc, links, identifier = "RL_ID")
head(toy_apdc, 10L)
```

Here, we have used the fact that if two episodes overlap, then one
episode starts during the other.

Some of the common types of links have been implemented as functions in
the `journeyer` package. For a complete list of these predicates, see
`?LinkPredicates`. To use these, note that you will need to use the
bang-bang operator `!!` from the tidyverse. For example, to find
transfers where both source and destination episode have evidence of a
transfer, or overlapping episodes, we could do:

```{r}
links <- pairs |>
  find_links(
    overlap = !!overlap_episodes(episode_start_parent, episode_end_parent, episode_start_child),
    transfer = !!transfers_both(
      episode_end_parent, mode_of_separation_recode_parent,
      facility_identifier_recode_parent, facility_trans_to_recode_parent,
      episode_start_child, source_of_referral_recode_child,
      facility_identifier_recode_child, facility_trans_from_recode_child
    )
  ) |>
  select(RL_ID_parent, RL_ID_child)

collect_journeys(toy_apdc, links, identifier = "RL_ID")
head(toy_apdc, 10L)
```

## Using `dbplyr` to work with large datasets

For large datasets, with many billions of pairs to check, a researcher
may reasonably be concerned about system load while doing this with base
R. Fortunately, the default predicates have been carefully constructed
to interface with `dbplyr`, allowing the work to be done through a
database management system with better resource management. For example,
using the `duckdb` package, an in-process OLAP system:

```{r}
library(duckdb)
library(dbplyr)

con <- dbConnect(duckdb(shared_home = FALSE), timezone_out = '', tz_out_convert = 'force')

# Tell DuckDB to read the data frame from R; DuckDB can also read Parquet files
duckdb_register(con, 'toy_apdc', toy_apdc)
toydata <- tbl(con, 'toy_apdc')

pairs <- inner_join(toydata, toydata, by = 'PPN', suffix = c('_parent', '_child')) |>
  filter(RL_ID_parent != RL_ID_child)

links <- pairs |>
  find_links(!!!default_predicates$NSW$APDC_APDC) |>
  select(RL_ID_parent, RL_ID_child) |>
  collect() # Tell dbplyr to run the query

toyids <- toydata |>
  select(RL_ID) |>
  collect()
setDT(toyids)

collect_journeys(toyids, links, identifier = 'RL_ID')
head(toyids, 10L)

on.exit(dbDisconnect(con, shutdown = TRUE), add = TRUE)
```

This table now contains the record identifier and the POHC number. It
can be inserted back into or registered with the database to conduct
further analyses.
